lifespan_eda

### 2023-12-11 eda/lifespan_eda.Rmd
adding log-transform for global volume features pre-combat (for lifespan analyses)

source("/Users/megardn/Desktop/BGD_Repos/GAMLSS/plotting_functions.R")
source("/Users/megardn/Desktop/BGD_Repos/combat_biovar/R_scripts/gamlss_helper_funs.R")
```

Load data from Jakob
```{r}
cn.df <- fread("data/lifespan_dataset_dkatlas_CNonly.csv", stringsAsFactors = TRUE)

@@

```{r}
fwrite(cn.df.site, file="data/lifespan_CN_site-level.csv")
```

@@

```{r}
fwrite(cn.df.imp, file="data/lifespan_CN_imp-sites.csv")
```

>>>>

vol_list_global <- readRDS(file=paste0(base_path, "/cubic_scripts/R_scripts/vol_list_global.rds"))

base_path <- "/Users/megardn/Desktop/BGD_Repos/combat_biovar"
source("/Users/megardn/Desktop/BGD_Repos/GAMLSS/plotting_functions.R")
source(paste0(base_path, "/cubic_scripts/R_scripts/gamlss_helper_funs.R"))
```

Load data from Jakob
```{r}
cn.df <- fread(paste0(base_path,"/data/lifespan_dataset_dkatlas_CNonly.csv"), stringsAsFactors = TRUE)

@@

```{r}
fwrite(cn.df.site, file=paste0(base_path, "/data/lifespan_CN_site-level.csv"))
```

looking into whether all global measures should be log-transformed to account for skew (getting neg vals for ventricles with combat-GAM)

```{r}
global_vols <- cn.df.site %>%
  dplyr::select(sex, log_age, all_of(vol_list_global)) %>%
  pivot_longer(cols=c(-sex, -log_age), names_to = "pheno", values_to = "volume") %>%
  mutate(log_vol = log(volume, base=10)) #test log-transform

ggplot(global_vols) +
  geom_point(aes(x=log_age, y=volume, color=sex), alpha=0.4) +
  facet_wrap(~pheno, scales="free_y")

ggplot(global_vols) +
  geom_histogram(aes(x=volume, fill=pheno), alpha=0.5) +
  facet_wrap(~pheno, scales="free")

ggplot(global_vols) +
  geom_histogram(aes(x=log_vol, fill=pheno), alpha=0.5) +
  facet_wrap(~pheno, scales="free")
```

@@

```{r}
fwrite(cn.df.imp, file=paste0(base_path, "/data/lifespan_CN_imp-sites.csv"))
```

## Visualize Harmonized Data

comparing ventricle vals across harmonization pipelines (neg vals in `lifespan-CN-imp-sites_cf.gam_data.csv`), may need to log-transform
```{r}
lifespan_imp_gam.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-imp-sites_cf.gam_data.csv"), stringsAsFactors = TRUE)
lifespan_imp_gamlss.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-imp-sites_cf.gamlss_data.csv"), stringsAsFactors = TRUE)

lifespan_gam.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-site-level_cf.gam_data.csv"), stringsAsFactors = TRUE)
lifespan_gamlss.df <- fread(paste0(base_path, "/data/lifespan/lifespan-CN-site-level_cf.gamlss_data.csv"), stringsAsFactors = TRUE)
```

```{r}
#mark negative vals
# cn.df.site <- as.data.frame(cn.df.site)
neg1 <- cn.df.site %>%
  dplyr::filter(Ventricles <= 0)
ven_raw <- ggplot() +
  geom_point(data=cn.df.site, aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg1, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("Raw Data")

neg2 <- lifespan_gam.df[lifespan_gam.df$Ventricles <=0]
ven_gam <- ggplot() +
  geom_point(data=lifespan_gam.df, aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg2, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBat-GAM")

neg3 <- lifespan_gamlss.df[lifespan_gamlss.df$Ventricles <=0]
ven_gamlss <- ggplot(lifespan_gamlss.df) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg3, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBatLS")

# cn.df.imp <- as.data.frame(cn.df.imp)
neg4 <- cn.df.imp %>%
  dplyr::filter(Ventricles <= 0)
ven_raw.imp <- ggplot(cn.df.imp) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg4, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("Raw Data - All Studies")

neg5 <- lifespan_imp_gam.df[lifespan_imp_gam.df$Ventricles <=0]
ven_gam.imp <- ggplot(lifespan_imp_gam.df) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg5, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBat-GAM - All Studies")

neg6 <- lifespan_imp_gamlss.df[lifespan_imp_gamlss.df$Ventricles <=0]
ven_gamlss.imp <- ggplot(lifespan_imp_gamlss.df) +
  geom_point(aes(x=log_age, y=Ventricles, color=sex), alpha =0.3) +
  geom_point(data=neg6, aes(x = log_age, y=Ventricles), color='red', alpha=0.6) + 
  ggtitle("ComBatLS - All Studies")

ggarrange(ven_raw, ven_gam, ven_gamlss, ven_raw.imp, ven_gam.imp, ven_gamlss.imp, common.legend = TRUE)
```
